---
title: "Group exercise 3: The spread of COVID-19 in the US"
author: "Emily and Sarah"
subtitle: 'DATA5207: Data Analysis in the Social Sciences'
output:
  html_document:
    toc: no
    df_print: paged
  pdf_document:
    toc: no
    latex_engine: xelatex
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, messages = FALSE)

library (tidyverse)
library(sf)
library(janitor)
library(sf)
library(scales) 
library(patchwork)
library(ggplot2)
library(rmapshaper)
library(tigris)
library(plotly)
devtools::install_github('bbc/bbplot')

```

# Load Data and Preprocess

```{r load google mobility data}
google.mobility <- read.csv('Data/Global_Mobility_Report.csv') %>%
  filter(country_region == 'United States')

```

```{r}
confirmed.cases.data <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

```

```{r load and clean confirmed case data by county}


# load packages

library(tidyr)
library(DataCombine)


# clean data

confirmed.cases.data2 <- confirmed.cases.data %>%
  dplyr::rename(state = Province_State,
                county = Admin2) %>% 
  dplyr::select(-UID, -iso2, -iso3, -code3,
                -Lat, -Long_, -Combined_Key, -Country_Region) %>% 
    gather(date, confirmed.cases,-state,-county,-FIPS) %>%
  dplyr::mutate(date = gsub('X', '', date), 
                date = as.Date(as.character(date), "%m.%d.%y"),
                county_state = paste0(county, ', ', state)) %>%
  arrange(county_state, date) %>%
  dplyr::mutate(lag = slide(.,
                                  Var = 'confirmed.cases', 
                                  NewVar = 'new',
                                  GroupVar = 'county_state', 
                                  slideBy = -1)[,'new'],
                new.cases = confirmed.cases - lag)


## calculate rolling averages and remove non-states

covid.smooth.data_county <- confirmed.cases.data2 %>%
    dplyr::group_by(county_state, date) %>% 
    dplyr::summarise(new.cases = sum(new.cases)) %>%
      dplyr::group_by(county_state) %>% 
    dplyr::mutate(cases_14days = zoo::rollmean(new.cases, 
                                               k = 14, fill = 0)) %>% 
  dplyr::ungroup() %>% 
  mutate() %>% 
  merge(confirmed.cases.data2 %>% 
          dplyr::select(county_state, date, county, state)) %>% 
  filter(!state %in% c('American Samoa', 
                       'Diamond Princess',
                       'Grand Princess',
                       'Guam',
                       'Northern Mariana Islands',
                       'Puerto Rico',
                       'Virgin Islands'))
```

```{r load additional predictor data}


load("Data/county.data.RData")

```

#### Climate DF

```{r}
# patterns for cleaning county names

patterns <- c(' Borough| Census Area| County| Parish| city')

climate.data <- county.data %>% 
    # grab file with full state names 
  
  left_join(read.csv('Data/fips concordance.csv') %>% 
              dplyr::select(State = Alpha.code, 
                            state_full = Name)) %>%
         dplyr::mutate(county = gsub(patterns, '', Area_Name),
                       county_state = paste0(county, ', ', state_full),
                       state = state_full) %>% 
    
    dplyr::select(county,
                  county_state,
                  state,
                  avg.dec.temp = Dec.Temp.AVG...F,
                  avg.jan.temp = Jan.Temp.AVG...F,
                  avg.feb.temp = Feb.Temp.AVG...F,
                  avg.mar.temp = Mar.Temp.AVG...F,
                  avg.apr.temp = Apr.Temp.AVG...F,
                  avg.may.temp = May.Temp.AVG...F,
                  avg.jun.temp = Jun.Temp.AVG...F,
                  avg.jul.temp = Jul.Temp.AVG...F,
                  avg.aug.temp = Aug.Temp.AVG...F,
                  avg.sep.temp = Sep.Temp.AVG...F,
                  avg.oct.temp = Oct.Temp.AVG...F,
                  avg.nov.temp = Nov.Temp.AVG...F)
    
```

#### Mobility DF

```{r}
mobility.data <- google.mobility %>% 
          dplyr::mutate(county = gsub(patterns, '', sub_region_2),
                        state = sub_region_1,
                        date = as.Date(date)) %>%
  dplyr::select(county,
                state,
                date,
                mobility.transit = transit_stations_percent_change_from_baseline,
                mobility.workplaces = workplaces_percent_change_from_baseline,
                mobility.grocery = grocery_and_pharmacy_percent_change_from_baseline)
  
```

```{r}
mobility.data <- na.omit(mobility.data)
```

```{r checking blank rows}
blank_counts <- sapply(mobility.data, function(col) sum(col == ""))
print(blank_counts)
```

```{r removing said blank rows}
clean.mobility.data <- mobility.data %>% filter(county != "")
clean.mobility.data <- clean.mobility.data %>% filter(state != "")
```

#### Transit DF

```{r}
# patterns for cleaning county names

patterns <- c(' Borough| Census Area| County| Parish| city')

transit.data <- county.data %>% 
    # grab file with full state names 
  
  left_join(read.csv('Data/fips concordance.csv') %>% 
              dplyr::select(State = Alpha.code, 
                            state_full = Name)) %>%
         dplyr::mutate(county = gsub(patterns, '', Area_Name),
                       county_state = paste0(county, ', ', state_full),
                       state = state_full) %>% 
    
    dplyr::select(county,
                  county_state,
                  state,
                  transit.scores = transit_scores...population.weighted.averages.aggregated.from.town.city.level.to.county) 
    
```

#### Collated Data

```{r recoding data}

# patterns for cleaning county names

patterns <- c(' Borough| Census Area| County| Parish| city')


# covid case data 
load("Data/covid.RData")

covid.data <- covid.smooth.data_county %>% 
  
  # county data

left_join(county.data %>% 
            
            # grab file with full state names 
  
  left_join(read.csv('Data/fips concordance.csv') %>% 
              dplyr::select(State = Alpha.code, 
                            state_full = Name)) %>%
         dplyr::mutate(county = gsub(patterns, '', Area_Name),
                       county_state = paste0(county, ', ', state_full)) %>% 
    
    dplyr::select(county_state,
                  FIPS = FIPS,
                  total.population = POP_ESTIMATE_2018,
                  pop.density = Density.per.square.mile.of.land.area...Population,
                  age.65.plus = Total_age65plus,
                  avg.dec.temp = Dec.Temp.AVG...F,
                  avg.jan.temp = Jan.Temp.AVG...F,
                  avg.feb.temp = Feb.Temp.AVG...F,
                  avg.mar.temp = Mar.Temp.AVG...F,
                  avg.apr.temp = Apr.Temp.AVG...F,
                  avg.may.temp = May.Temp.AVG...F,
                  avg.jun.temp = Jun.Temp.AVG...F,
                  avg.jul.temp = Jul.Temp.AVG...F,
                  avg.aug.temp = Aug.Temp.AVG...F,
                  avg.sep.temp = Sep.Temp.AVG...F,
                  avg.oct.temp = Oct.Temp.AVG...F,
                  avg.nov.temp = Nov.Temp.AVG...F,
                  icu.beds = ICU.Beds,
                  active.physicians = Active.Physicians.per.100000.Population.2018..AAMC.,
                  active.patient.care.physicians = Total.Active.Patient.Care.Physicians.per.100000.Population.2018..AAMC.,
                  housing.density = Density.per.square.mile.of.land.area...Housing.units,
                  transit.scores = transit_scores...population.weighted.averages.aggregated.from.town.city.level.to.county)) %>%
    
    
  merge(google.mobility %>% 
          dplyr::mutate(county = gsub(patterns, '', sub_region_2),
                        county_state = paste0(county, ', ', sub_region_1),
                        date = as.Date(date),
                        mobility.transit = ifelse(is.na(transit_stations_percent_change_from_baseline), 0,
                                   transit_stations_percent_change_from_baseline)) %>% 
  
  dplyr::select(county_state,
                date,
                mobility.transit,
                mobility.workplaces = workplaces_percent_change_from_baseline,
                mobility.grocery = grocery_and_pharmacy_percent_change_from_baseline)) %>% 
  
  # convert to rates
  
  mutate(new.cases = new.cases / (total.population / 100000),
  #        icu.beds = icu.beds / (total.population / 100000),
         age.65.plus = age.65.plus / total.population)
  #        avg.winter.temp = (avg.dec.temp + avg.jan.temp + avg.feb.temp) / 3)


# lag new cases

new.cases.rate.lagged <- DataCombine::slide(covid.data,
                                     Var = 'new.cases', 
                                     NewVar = 'new.cases.lag',
                                     GroupVar = 'county_state', 
                                     slideBy = -1) 

# lag workplace mobility 

workplace.lag14 <- DataCombine::slide(covid.data,
                            Var = 'mobility.workplaces', 
                            NewVar = 'workplaces.lag14',
                            GroupVar = 'county_state', 
                            slideBy = -14)


# combine 

covid.data <- covid.data %>% 
  
  merge(new.cases.rate.lagged %>% 
            dplyr::select(county_state, date, 
                        new.cases.lag)) %>% 
  
    merge(workplace.lag14 %>% 
            dplyr::select(county_state, date, 
                        workplaces.lag14))


#covid.data <- covid.data %>%
 # merge(confirmed.cases.data %>%
  #        dplyr::rename(state = Province_State,
   #                     county = Admin2) %>%
    #      dplyr::select(-Lat, -Long_))

```

```{r}
colnames(covid.data)
```

Add shape files for any mapping tasks later

```{r link us data with shape files, warning=FALSE, message=FALSE}

us.shape.1 <- 
  st_read("Data/US shapefiles/cb_2016_us_county_20m.shp")
us.shape.1 <- us.shape.1 %>%
  shift_geometry() #This function moves alaska into a easier area to interpret

```

```{r}

fip.concordance <- read.csv("Data/fips concordance.csv")

us.shape.2 <- us.shape.1 %>% 
  merge(fip.concordance %>%
          mutate(STATEFP =  ifelse(Numeric.code >= 0 & Numeric.code <= 9, 
                             paste0(0 , Numeric.code), Numeric.code)) %>%
          dplyr::rename(state = Alpha.code))


us.shape.2$county <- paste0(us.shape.2$NAME, ", ", us.shape.2$state)

us.shape.2$county <- gsub("Doña Ana, NM", "Dona Ana, NM", us.shape.2$county)
us.shape.2$county <- gsub("LaSalle, LA", "La Salle, LA", us.shape.2$county)
us.shape.2$county <- gsub("Oglala Lakota, SD", "Oglala, SD", us.shape.2$county)

```

```{r link shapefiles with fip concordance file}

# Summary statistics for age over 65
summary(covid.data$age.65.plus)

# Correlation between age over 65 and new cases
cor(covid.data$age.65.plus, covid.data$new.cases, use = "complete.obs")

# Scatter plot
plot(covid.data$new.cases, covid.data$age.65.plus, ylab = "Age Over 65", xlab = "New COVID-19 Cases", main = "Age Over 65 vs. New COVID-19 Cases")


```

```{r}

specific_date <- as.Date("2020-07-20	")  
filtered_data <- covid.data %>%
  filter(date == specific_date)
# Summary statistics for population density
summary(covid.data$pop.density)

# Correlation between population density and new cases
cor(covid.data$pop.density, covid.data$new.cases, use = "complete.obs")

# Scatter plot
plot(filtered_data$new.cases, filtered_data$pop.density, ylab = "Population Density", xlab = "New COVID-19 Cases", main = "Population Density vs. New COVID-19 Cases")
```

```{r}
# Summary statistics for active physicians per 100,000
summary(covid.data$active.patient.care.physicians)

# Correlation between active physicians and new cases
cor(covid.data$active.patient.care.physicians, covid.data$new.cases, use = "complete.obs")

# Scatter plot
plot(covid.data$new.cases, covid.data$active.patient.care.physicians, ylab = "Active Physicians per 100,000", xlab = "New COVID-19 Cases", main = "Physicians vs. New COVID-19 Cases")
```

# Descriptive Analysis

## Transit

| Transit Score® | Description                         |
|----------------|-------------------------------------|
| 90–100         | Rider's Paradise                    |
|                | World-class public transportation.  |
| 70–89          | Excellent Transit                   |
|                | Transit is convenient for most trips.|
| 50–69          | Good Transit                        |
|                | Many nearby public transportation options. |
| 25–49          | Some Transit                        |
|                | A few nearby public transportation options. |
| 0–24           | Minimal Transit                     |
|                | It is possible to get on a bus.     |

```{r}

na_count_per_column <- transit.data %>%
  summarise_all(~ sum(is.na(.)))
print(na_count_per_column)

```

```{r remove NA values}
clean.transit.data <- na.omit(transit.data)

# standardise
clean.transit.data <- clean.transit.data %>%
  mutate(z.transit = (transit.scores - mean(transit.scores, na.rm=T)) / 
           sd(transit.scores, na.rm=T))
```

```{r}
ggplot(clean.transit.data, aes(x = z.transit)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Standardised transit scores in America",
       x = "Transit Scores",
       y = "Frequency") +
  bbplot::bbc_style()+
  theme(plot.title = element_text(size = 12),
        axis.title.x = element_text(size = 8), # Adjust the x-axis label size
    axis.title.y = element_text(size = 8))
```

```{r link the geometry with transit data}
us.shape.2 <- us.shape.2 %>%
  select(-county, -Name, -Status, -COUNTYFP, -COUNTYNS, -STATEFP, -AFFGEOID, -LSAD, -ALAND, -AWATER, -Numeric.code, -Status)%>%
  rename(county = NAME)

transit.map <- full_join(us.shape.2,
                         clean.transit.data %>% 
                           select(county, state, z.transit), 
                         by = "county") %>%
              st_as_sf()
```

```{r}

ggplot(transit.map) +
  geom_sf(mapping = aes(fill = z.transit), colour = NA, inherit.aes = FALSE) +
  scale_fill_gradient2(
    low = "#2c7bb6",           # Blue
    mid = "#ffffbf",           # Yellow (midpoint)
    high = "#d7191c",          # Red
    midpoint = mean(transit.map$z.transit, na.rm = TRUE),
    limits = c(min(transit.map$z.transit, na.rm = TRUE),
               max(transit.map$z.transit, na.rm = TRUE)),
    label = scales::comma,
    na.value = "grey"
  ) +
  labs(title = 'Standardised Transit Distribution') +
  guides(fill = guide_legend(keyheight = 0.3)) +
  theme_void() +
  bbplot::bbc_style() +
  theme(
    plot.title = element_text(size = 16),         # Make the title larger
    legend.position = 'bottom',
    legend.title = element_blank(),
    legend.text = element_text(size = 10),        # Make legend text smaller
    axis.title = element_text(size = 10),         # Make axis labels smaller
    axis.text = element_text(size = 8)            # Make axis text smaller
  )

```

## Mobility

```{r}
# install.packages("zoo")
library(zoo)
```

```{r summarise data}
summarise.mobility.data <- clean.mobility.data %>%
  group_by(date) %>%
  summarise(mobility.transit = mean(mobility.transit, na.rm = TRUE),
            mobility.grocery = mean(mobility.grocery, na.rm = TRUE),
            mobility.workplaces = mean(mobility.workplaces, na.rm = TRUE))


# Mobility Values: Each value in the dataset represents a percentage change from a baseline.
# Rolling Average: The 7-day rolling average smooths these daily percentage changes to show a trend over time.
# Percentage Representation: The rolling average remains a percentage because it’s derived from percentage values.

# Calculate rolling 7 day average
grocery.mobility.data <- summarise.mobility.data %>%
  arrange(date) %>%
  mutate(rolling_avg = zoo::rollmean(mobility.grocery, 7, fill = NA, align = "right"))

transit.mobility.data <- summarise.mobility.data %>%
  arrange(date) %>%
  mutate(rolling_avg = zoo::rollmean(mobility.transit, 7, fill = NA, align = "right"))


workplace.mobility.data <- summarise.mobility.data %>%
  arrange(date) %>%
  mutate(rolling_avg = zoo::rollmean(mobility.workplaces, 7, fill = NA, align = "right"))


```

```{r}
ggplot(grocery.mobility.data, aes(x = date, y = rolling_avg)) +
  geom_line(color = "blue") +
  labs(title = "Changes in Visitors to Grocery and Pharmacy Stores Over Time",
       x = "Date",
       y = "7-day Rolling Average of Mobility (%)") +
  bbplot::bbc_style()+
  theme(plot.title = element_text(size = 12),
        axis.title.x = element_text(size = 8), # Adjust the x-axis label size
    axis.title.y = element_text(size = 8))
```

```{r}
ggplot(grocery.mobility.data, aes(x = date, y = rolling_avg)) +
  geom_line(color = "blue", size = 1) +  # Thicker line for better visibility
  # geom_point(color = "red", size = 1, alpha = 0.7) +  # Adding points for data
  labs(title = "Changes in Visitors to Grocery and Pharmacy Stores Over Time",
       x = "Date",
       y = "7-day Rolling Average of Mobility (%)") +
  bbplot::bbc_style() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),  # Larger, bold title
    axis.title.x = element_text(size = 12),  # Larger x-axis label
    axis.title.y = element_text(size = 12),  # Larger y-axis label
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    axis.text.y = element_text(size = 10)
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +  # Format x-axis dates
  geom_vline(xintercept = as.numeric(as.Date(c("2020-03-11", "2020-12-25"))),  # Add vertical lines for significant dates
             linetype = "dashed", color = "grey50") +
  annotate("text", x = as.Date("2020-03-11"), y = max(grocery.mobility.data$rolling_avg, na.rm = TRUE)* -0.5,
           label = "Pandemic Declared", angle = 90, vjust = -0.5, size = 3, color = "grey50") +
  annotate("text", x = as.Date("2020-12-25"), y = max(grocery.mobility.data$rolling_avg, na.rm = TRUE)* 0.8,
           label = "Christmas", angle = 90, vjust = -0.5, size = 3, color = "grey50")
```

```{r}
ggplot(transit.mobility.data, aes(x = date, y = rolling_avg)) +
  geom_line(color = "blue") +
  labs(title = "Changes in Visitors to Transit Over Time",
       x = "Date",
       y = "7-day Rolling Average of Mobility (%)") +
  theme_minimal()
```

```{r}
ggplot(workplace.mobility.data, aes(x = date, y = rolling_avg)) +
  geom_line(color = "blue") +
  labs(title = "Changes in Visitors to Workplaces Over Time",
       x = "Date",
       y = "7-day Rolling Average of Mobility (%)") +
  theme_minimal()
```

```{r}


# Add a type column to each dataframe
grocery.mobility.data <- grocery.mobility.data %>%
  mutate(type = "Grocery and Pharmacy")

transit.mobility.data <- transit.mobility.data %>%
  mutate(type = "Transit")

workplace.mobility.data <- workplace.mobility.data %>%
  mutate(type = "Workplace")

# Combine the dataframes into one
combined_mobility_data <- bind_rows(grocery.mobility.data, transit.mobility.data, workplace.mobility.data)

# Create the ggplot
p <- ggplot(combined_mobility_data, aes(x = date, y = rolling_avg, color = type)) +
  geom_line() +
  labs(title = "Changes in Mobility Over Time",
       x = "Date",
       y = "7-day Rolling Average of Mobility (%)",
       color = "Mobility Type") +
  theme_minimal()

# Convert the ggplot to an interactive plotly plot
interactive_plot <- ggplotly(p)

# Display the interactive plot
interactive_plot

```

## Climate data

```{r}
# Changing the climate data from a wide format to long format:
climate.data <- climate.data %>%
  dplyr::select(county, 
                state,
                avg.temp.jan = avg.jan.temp,
                avg.temp.feb = avg.feb.temp,
                avg.temp.mar = avg.mar.temp,
                avg.temp.apr = avg.apr.temp,
                avg.temp.may = avg.may.temp,
                avg.temp.jun = avg.jun.temp,
                avg.temp.jul = avg.jul.temp,
                avg.temp.aug = avg.aug.temp,
                avg.temp.sep = avg.sep.temp,
                avg.temp.oct = avg.oct.temp,
                avg.temp.nov = avg.nov.temp,
                avg.temp.dec = avg.dec.temp)


climate.data.long <- climate.data %>%
  pivot_longer(
    cols = starts_with("avg.temp."),
    names_to = "month",
    values_to = "temp"
  )

climate.data.long <- climate.data.long %>%
  mutate(month = case_when(
    month == "avg.temp.jan" ~ "Jan",
    month == "avg.temp.feb" ~ "Feb",
    month == "avg.temp.mar" ~ "Mar",
    month == "avg.temp.apr" ~ "Apr",
    month == "avg.temp.may" ~ "May",
    month == "avg.temp.jun" ~ "Jun",
    month == "avg.temp.jul" ~ "Jul",
    month == "avg.temp.aug" ~ "Aug",
    month == "avg.temp.sep" ~ "Sep",
    month == "avg.temp.oct" ~ "Oct",
    month == "avg.temp.nov" ~ "Nov",
    month == "avg.temp.dec" ~ "Dec",
    TRUE ~ month
  ))
```

```{r link us data with shape files2, message=FALSE}
# us.shape.2 <- us.shape.2 %>%
#   select(-county, -Name, -Status, -COUNTYFP, -COUNTYNS, -STATEFP, -AFFGEOID, -GEOID, -LSAD, -ALAND, -AWATER, -Numeric.code, -Status)%>%
#   rename(county = NAME)

# Don't want to include missing data since its not useful - hence inner join 
climate.map <- full_join(us.shape.2,
                         climate.data.long %>% 
                           select(county, state, month, temp), 
                         by = "county") %>%
              st_as_sf()
 # To acknolwedge its a shape file
```

```{r collate climate data and plot over time}

climate.state <- climate.data.long %>%
  group_by(state, month)%>%
  summarise(avg_temp = mean(temp, na.rm= TRUE, .groups = 'drop'))


ggplot(climate.state, aes(x = month, y = avg_temp, group = state, color = state)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Temperature Over Each Month Per State",
       x = "Month",
       y = "Average Temperature") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),  # Set title size
    axis.title.x = element_text(size = 10), # Set x-axis label size
    axis.title.y = element_text(size = 10), # Set y-axis label size
    legend.position = "bottom",  # Move legend below plot
    legend.title = element_blank(),  # Remove legend title
    legend.text = element_text(size = 8)  # Set legend text size
  ) +
  guides(color = guide_legend(nrow = 5))  # Set legend to one row
```

```{r}


# Create the plot using ggplot
p <- ggplot(climate.state, aes(x = month, y = avg_temp, group = state, color = state)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Temperature Over Each Month Per State",
       x = "Month",
       y = "Average Temperature") +
  theme_minimal()

# Convert the ggplot object to plotly
p <- ggplotly(p, tooltip = c("x", "y", "color"))

# Show the interactive plot
p
```

```{r map the average temperature for each month in each county}
month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Convert the month variable to a factor with the desired order
climate.map$month <- factor(climate.map$month, levels = month_order)

ggplot(climate.map) +
  geom_sf(mapping = aes(fill = temp), colour = NA, inherit.aes = FALSE) +
  scale_fill_gradient2(low = "#0072B2",
                       high = "red",
                       midpoint = mean(climate.map$temp, na.rm = TRUE),
                       limits = c(min(climate.map$temp, na.rm = TRUE),
                                  max(climate.map$temp, na.rm = TRUE)),
                       label = scales::comma,
                       na.value = "grey") +
  labs(title = 'Average Climate distribution') +
  guides(fill = guide_legend(keyheight = 0.3)) +
  theme_void() +
  theme(plot.title = element_text(size = 7),
        legend.position = 'bottom',
        legend.title = element_blank()) +
  facet_wrap(~month) # This line tells ggplot to create separate panels for each unique value in the month column. Each panel will display the map for one month.

```

# Sarah's stuff :)

```{r}

# Ensure FIPS column is correctly named and of the same type in both data frames
county.data <- county.data %>%
  mutate(FIPS = as.character(FIPS))

us.shape.2 <- us.shape.2 %>%
  mutate(GEOID = as.character(GEOID))

covid.data <- covid.data %>%
  mutate(FIPS = as.character(FIPS))
```

```{r}

county.data$FIPS <- str_pad(as.character(county.data$FIPS), width = 5, pad = "0")

# Function to perform join in chunks
join_in_chunks <- function(shape_data, covid_data, chunk_size = 1000) {
  n <- nrow(shape_data)
  result <- NULL
  
  for (i in seq(1, n, by = chunk_size)) {
    chunk <- shape_data[i:min(i + chunk_size - 1, n), ]
    chunk_result <- chunk %>%
      left_join(covid_data, by = c("GEOID" = "FIPS"))
    
    if (is.null(result)) {
      result <- chunk_result
    } else {
      result <- bind_rows(result, chunk_result)
    }
  }
  
  return(result)
}

# Perform join in chunks


county_data_geo <- join_in_chunks(us.shape.2, county.data, chunk_size = 1000)
```

```{r}
county.covid = join_in_chunks(us.shape.2, covid.data, chunk_size = 1000)
```


```{r}
ggplot(data = county_data_geo) +
      geom_sf(aes(fill = log(Density.per.square.mile.of.land.area...Population)), color = NA) +
      scale_fill_viridis_c(option = "viridis") +
      theme_minimal() +
      labs(title = "Population Density by County Standardised", 
           fill = "Density (per sq. mile)")

```

```{r}
ggplot(data = county_data_geo) +
  geom_sf(aes(fill = log(Total_age85plusr/POP_ESTIMATE_2018)), color = NA) +
  scale_fill_viridis_c(option = "viridis") +
  theme_minimal() +
  labs(title = "Proportion of Population Age 65+ by County Standardised",
       fill = "Proportion")
```


```{r}
# Plot Active Physicians per 100,000
ggplot(data = county_data_geo) +
  geom_sf(aes(fill = log(Total.Active.Patient.Care.Physicians.per.100000.Population.2018..AAMC.)), color = NA) +
  scale_fill_viridis_c(option = "viridis") +
  theme_minimal() +
  labs(title = "Active Physicians per 100,000 by County Standardised",
       fill = "Physicians/100k")
```

```{r}

county.covid$date <- as.Date(county.covid$date)

# Filter the data for the year 2021
county.covid_2021 <- county.covid %>%
  filter(year(date) == 2021)
```

```{r}
# Ensure the geometry column is in the correct format and extract centroids
county.covid_2021 <- st_transform(county.covid_2021, 4326)
county.covid_2021$centroid <- st_centroid(st_geometry(county.covid_2021))

# Extract coordinates from centroids
county.covid_df <- county.covid_2021 %>%
  mutate(long = st_coordinates(centroid)[,1],
         lat = st_coordinates(centroid)[,2]) %>%
  as.data.frame() %>%
  select(-geometry, -centroid)  # Remove geometry columns as they are now in long and lat

```



```{r}
start_date <- as.Date("2021-01-01")
filtered_dates <- seq(start_date, max(county.covid_df$date), by = "14 days")
county.covid_df <- county.covid_df %>%
  filter(date %in% filtered_dates)

# Prepare data for plotly
county.covid_long <- county.covid_df %>%
  gather(key = "variable", value = "value", -GEOID, -date, -cases_14days, -long, -lat)

# Check the resulting data
head(county.covid_long)
```


```{r}
# Create the plot using plotly
fig <- plot_ly(county.covid_long, 
               x = ~long, 
               y = ~lat, 
               frame = ~date, 
               ids = ~GEOID, 
               type = 'scatter', 
               mode = 'markers', 
               marker = list(color = ~cases_14days, colorscale = 'Viridis', showscale = TRUE)) %>%
  layout(title = "New COVID-19 Cases by County in 2021 Fortnightly Average",
         xaxis = list(title = "Longitude"),
         yaxis = list(title = "Latitude")) %>%
  animation_opts(frame = 1000, easing = "elastic", redraw = TRUE) %>%
  animation_button(x = 1, xanchor = "right", y = 0, yanchor = "bottom")

# Render the plotly object
## fig
```


```{r}
# Create a simpler plot without additional options
fig_simple <- plot_ly(county.covid_long,
                      x = ~long,
                      y = ~lat,
                      frame = ~date,
                      ids = ~GEOID,
                      type = 'scattergl',  # Use scattergl for WebGL rendering
                      mode = 'markers')

# Render the simpler plotly object
fig_simple
```






# Model

"county_state" "date" "new.cases"\
[4] "cases_14days" "county" "state"\
[7] "FIPS" "total.population" "pop.density"\
[10] "age.65.plus" "avg.dec.temp" "avg.jan.temp"\
[13] "avg.feb.temp" "avg.mar.temp" "avg.apr.temp"\
[16] "avg.may.temp" "avg.jun.temp" "avg.jul.temp"\
[19] "avg.aug.temp" "avg.sep.temp" "avg.oct.temp"\
[22] "avg.nov.temp" "icu.beds" "active.physicians"\
[25] "active.patient.care.physicians" "housing.density" "transit.scores"\
[28] "mobility.transit" "mobility.workplaces" "mobility.grocery"\
[31] "new.cases.lag" "workplaces.lag14"

```{r see the significance of temperature}
model2 <- lm(cases_14days ~ avg.jan.temp + avg.feb.temp + avg.mar.temp + avg.apr.temp + avg.may.temp + avg.jun.temp + avg.jul.temp + avg.aug.temp + avg.sep.temp + avg.oct.temp + avg.nov.temp + avg.dec.temp , covid.data)

summary(model2)
```

```{r see the significance of mobility}
model3 <- lm(cases_14days ~mobility.transit +mobility.grocery +mobility.workplaces + avg.jan.temp + avg.feb.temp + avg.mar.temp + avg.apr.temp + avg.may.temp + avg.jun.temp + avg.jul.temp + avg.aug.temp + avg.sep.temp + avg.oct.temp + avg.nov.temp + avg.dec.temp, covid.data)
summary(model3)
```

```{r testing independence}
library(lmtest)

# Durbin-Watson test
dwtest(model3)

```

```{r testing homoskedasticity}
# Plot residuals vs fitted values
plot(fitted(model3), residuals(model3), main = "Residuals vs Fitted")
abline(h = 0, col = "red")
```

```{r}
# Q-Q plot
qqnorm(residuals(model3))
qqline(residuals(model3), col = "red")

```

# Results and data journalism project explaining the spread of COVID-19 in the US. Why has it impacted some

areas more than others?

COVID-19
